Local Search Approximation Algorithms for Partitional and Hierarchical Clustering

### 2015-05-10 tex/project.tex
Flat results and Hier Implementation

We implemented the local search swapping algorithm described above for $p=1, 2, 3$ as well as the supermodular function minimization algorithm in Python, and compare against $k$-means, $k$-medoids and Shi and Meila's spectral algorithm (implemented in the scikit-learn package for Python).  We also implemented the $k$-medoids algorithm with a maximum iteration of 1000. Note that for $p=1$, the multiswaps algorithm is equivalent to the well-known PAM algorithm.  In order to calculate approximation ratios, Gurobi was used to compute the global optimum for each data set.

@@

For each data set and each clustering method, we report objective function values, the approximation ratio, computation time, and adjusted mutual information (AMI).  AMI is a measure which scores accuracy of predicted clusters compared to a true labeling by measuring the agreement of two assignments. The measure is normalized to reflect chance, so that a perfect labeling receives a score of 1 and an independent random labeling receives a score of 0 in expectation.  See \cite{Vinh} for further information.  The results reported are average values across 10 trials.

@@

Comment on results here.

@@

\section{Conclusion}

>>>>

We implemented the local search swapping algorithm described above for $p=1, 2, 3$ as well as the supermodular function minimization algorithm in Python and compare against $k$-means, $k$-medoids and Shi and Meila's spectral algorithm (implemented in the scikit-learn package for Python).  We also implemented the $k$-medoids algorithm with a maximum iteration of 1000. Note that for $p=1$, the multiswaps algorithm is equivalent to the well-known PAM algorithm.  In order to calculate approximation ratios, Gurobi was used to compute the global optimum for each data set.

@@

For each data set and each clustering method, we report the approximation ratio, computation time, and adjusted mutual information (AMI).  AMI is a measure which scores accuracy of predicted clusters compared to a true labeling by measuring the agreement of two assignments. The measure is normalized to reflect chance, so that a perfect labeling receives a score of 1 and an independent random labeling receives a score of 0 in expectation.  See \cite{Vinh} for further information.  The results reported are average values across 10 trials.

@@

In terms of computation time, our proposed algorithms were much slower than $k$-means, the spectral algorithm, or $k$-medoids. In general, the multiswaps algorithm was around two orders of magnitude slower with a small multiplicative increase as $p$ increased, and the supermodular algorithm was around four orders of magnitude slower. This was probably due to the extra sampling the algorithm had to do.

In addition, the supermodular function did not perform as well in terms of AMI or approximation ratio compared to the other methods. Calculating the curvature, we saw that every  test set's curvature was almost equal to 1 so the guarantee of the algorithm was quite poor. Given that, the algorithm did maintain an approximation ratio below 2 for all instances and an AMI measure above 0.5 for almost all instances even averaging around 0.85 for the 7 dimensional Gaussian data.

When increasing $p$ for the multiswaps algorithm, we did see a slight increase in time but the times remained on the same order of magnitude. The increase in $p$ did not seem to have any significant effect on the AMI measure or approximation ratio with the values being roughly equivalent. This may due to the fact the algorithm for $p=1$ performed quite well already with an approximation ratio very close to 1 that beat out $k$-medoids. In addition, PAM had the highest AMI measures for the iris and soybean data sets. The AMI measures for the Gaussian data set were higher than those of $k$-medoids in general but equivalent or only slightly weaker than those of the spectral method or $k$-means. Adding in a few rounds of $k$-medoids at the end of our multiswaps algorithm did not seem to have much of an effect on the AMI measure but did boost the approximation ratio, sometimes finding an optimal solution.

It is interesting to note that higher dimensional Gaussian data seemed easier for the algorithms to classify along with smaller values of $k$ and lower percentage of noisy data, which makes sense based on our intuition of clustering. For the OR library instances, we saw that the approximation ratio tended to increase as $k$ grew larger but not as $n$ grew.

Overall, the results of our experiments do not point to the fact that these proposed approximation algorithms can add any advantage over current methods. However, the results do point to the fact that we can use sampling to run PAM with lower computational cost while still maintaining good performance accuracy both in terms of AMI measure and approximation ratios. Adding in a few rounds of $k$-medoids seems also to be a good method to refine a given solution towards a local optimum.

@@

\subsection{Implementation and Results}
We implemented the nesting hierarchical framework above in Python offering options for both the randomized and deterministic versions. As a black-box $k$-median solver, we used $k$-medoids and the 1-swap+ algorithms above. We compared the results to hierarchical agglomerative clustering with both average linkage and Ward's linkage (implemented in the scipy package in Python).

We also tested a more computationally friendly heuristic of the hierarchical framework which just buckets solutions for values of $k$ that are a power of two. This yields many fewer $k$-median instances to solve and reduced the run-time significantly. This was motivated by the fact that with the difference between various $k$-median solutions will be larger when $k$ is small so we expect our sampling to still represent much of the underlying structure.

For each data set and each clustering method, we report the approximation ratio, computation time, and FScore measure.  FScore is a measure that looks at the clusters represented on each level and how they represent the true clustering in terms of precision and recall. See \cite{Larsen} for more information.

\section{Conclusion and Future Work}
